---
title: "Overrepresentation Analysis with clusterProfiler: several molecular signatures in a loop"
format: html
editor: visual
---

This document runs the same Overrepresentation Analysis as `ora_clusterProfiler.qmd`, but for several molecular signatures (collections) at once. It is standalone: it loads the data and defines the DEGs itself. The results table and figures for each comparison and collection are saved in their own folder inside `output_loop/`, named comparison_collection (e.g. `output_loop/AD_vs_ctrol_male_C2KEGG/`), so they never overwrite the results of `ora_clusterProfiler.qmd` in `output/`.

The settings below (DEG cutoffs, background, `padj_cutoff`, `min_geneset_size`) are the same as in `ora_clusterProfiler.qmd`. If you change them in one document, change them in the other too.

### Install the packages, if you don't have them already

```{r, results="hide"}
# install.packages("BiocManager")

# Bioconductor packages
#bioc_packages <- c("clusterProfiler", "enrichplot", "ggupset", "msigdbr")

#BiocManager::install(setdiff(bioc_packages, rownames(installed.packages())), ask = FALSE)


# CRAN packages
#cran_packages <- c("tidyverse", "here")

#install.packages(setdiff(cran_packages, rownames(installed.packages())))
```

### Load the required libraries

```{r, results="hide"}
library(tidyverse)        # data manipulation and visualization (ggplot2, dplyr, tidyr, etc.)
library(clusterProfiler)  # enrichment analysis (ORA and GSEA)
library(enrichplot)       # visualization of enrichment analysis results
library(msigdbr)          # molecular signatures database (pathways, biological functions, etc.)
library(here)             # builds file paths from the project root (the folder containing ORA.Rproj)
```

### Set the path to your folder

```{r, results="hide"}
## Set relevant paths
base_path <- here() # root of the project (the folder containing ORA.Rproj)
make_path <- function(folder) {
  path <- file.path(base_path, folder) # build the full path by appending folder name to base path
  if (!dir.exists(path)) dir.create(path, recursive = TRUE) # create the folder if it doesn't exist
  path                                      # return the path
}

in_path  <- make_path("input")       # create and store path to the input folder
out_path <- make_path("output_loop") # separate output folder, so results from ora_clusterProfiler.qmd are not overwritten
```

### Upload the results of the DESeq2 analysis to the input folder

```{r}
df <- read_csv(file.path(in_path, "AD_vs_ctrol_male.csv"))
head(df)
```

### Settings the variables

```{r}
name_of_comparison  <- 'AD_vs_ctrol_male' ## label for the comparison, used for folder and file names
padj_cutoff         <- 0.05 # significance threshold for enriched pathways
min_geneset_size    <- 7  # the minimum size of the gene set within the universe
```

### Define the DEGs cutoffs (e.g., log2FoldChange, padj)

```{r}
df_sig <- df %>%
  filter(!is.na(padj), !is.na(log2FoldChange)) %>% # remove rows with missing padj or log2FoldChange values
  mutate(diffexpressed = case_when( # create column diffexpress and classfies according to log2FC value
    log2FoldChange > 0.5 & padj < 0.05 ~ "UP",
    log2FoldChange < -0.5 & padj < 0.05 ~ "DOWN",
    TRUE ~ "NO" # all remaining genes: not significantly differentially expressed
  )) %>%
  filter(diffexpressed != "NO") %>% # keep only significantly UP or DOWN regulated genes
  filter(!is.na(symbol)) %>% # remove genes without a gene symbol
  arrange(padj) %>% # sort by padj so the most significant row comes first
  distinct(symbol, .keep_all = TRUE) # remove duplicate gene symbols, keeping the most significant row

# Report, sanity check
cat("Found", nrow(df_sig), "DEGs\n")
cat("UP:  ", sum(df_sig$diffexpressed == "UP"), "\n")
cat("DOWN:", sum(df_sig$diffexpressed == "DOWN"), "\n")

# named vector of log2FC, used to color the heatplot and the gene frequency barplot by direction
gene_fc <- setNames(df_sig$log2FoldChange, df_sig$symbol)
```

### Prepare background gene list

```{r}
# extracting all unique, non-missing gene symbols from the full dataset to use as the statistical background (universe) for ORA
background_genes <- unique(na.omit(df$symbol))

cat("Number of background genes:", length(background_genes), "\n")
```

### Define the collections to run

Use `msigdbr_collections()` to find the collection and subcollection codes. `keywords` filters the pathways by name before plotting; `NA` means no keyword filter. `label` is used for the folder and file names, so give each row a different label.

```{r}
# msigdbr_collections() # uncomment to see the available collections and subcollections

collections <- tribble(
  ~label,       ~collection, ~subcollection,    ~keywords,
  "C2KEGG",     "C2",        "CP:KEGG_MEDICUS", NA, # use "CP:KEGG_LEGACY" for the classic KEGG gene sets
  "C2Reactome", "C2",        "CP:REACTOME",     NA,
  "C5BP",       "C5",        "GO:BP",           NA,
  "C7immune",   "C7",        "IMMUNESIGDB",     NA
)
```

### Define the functions used in the loop

```{r}
# theme shared by the barplot and dotplot
my_theme <- theme(
  axis.text.y = element_text(size = 6),
  axis.text.x = element_text(size = 8),
  plot.title  = element_text(size = 12),
  legend.text = element_text(size = 8)
)

# Saves a plot to PDF; if the plot fails (e.g. too few pathways), prints a message and keeps going
save_plot <- function(make_plot, file, width, height) {
  tryCatch({
    p <- suppressWarnings(make_plot())
    suppressWarnings(ggsave(filename = file, plot = p, width = width, height = height))
  }, error = function(e) message("  Skipped ", basename(file), ": ", conditionMessage(e)))
}

# Gene frequency barplot: number of filtered pathways each gene appears in, colored by the gene's log2 fold change
plot_gene_frequency <- function(res_filtered, title, n_top_genes = 30) {
  gene_freq <- as.data.frame(res_filtered) %>%
    dplyr::select(ID, geneID) %>%
    tidyr::separate_longer_delim(geneID, delim = "/") %>% # one row per gene per pathway
    dplyr::count(geneID, name = "n_pathways") %>%         # number of pathways each gene appears in
    dplyr::mutate(log2FoldChange = gene_fc[geneID]) %>%
    dplyr::arrange(desc(n_pathways)) %>%
    dplyr::slice_head(n = n_top_genes)

  fc_limit <- max(abs(gene_freq$log2FoldChange)) # symmetric color scale so 0 is always the neutral midpoint

  ggplot(gene_freq,
         aes(x = reorder(geneID, -n_pathways), y = n_pathways, fill = log2FoldChange)) +
    geom_col(width = 0.8) +
    scale_fill_gradient2(low = "#2166AC", mid = "grey90", high = "#B2182B", # blue = DOWN, red = UP
                         midpoint = 0, limits = c(-fc_limit, fc_limit)) +
    labs(x = NULL, y = "Number of pathways", title = title) +
    theme_minimal() +
    theme(axis.text.x = element_text(size = 7, angle = 90, hjust = 1, vjust = 0.5),
          panel.grid.major.x = element_blank(),
          plot.title  = element_text(size = 12))
}

# Runs ORA for one collection, saves the results table and the figures
run_collection <- function(label, collection, subcollection, keywords) {
  cat("\n---", label, "---\n")
  prefix    <- paste0(name_of_comparison, "_", label)
  coll_path <- file.path(out_path, prefix) # one folder per comparison and collection, e.g. output_loop/AD_vs_ctrol_male_C2KEGG
  if (!dir.exists(coll_path)) dir.create(coll_path, recursive = TRUE)
  # remove figures from a previous run so old plots are not mistaken for new results
  loop_plots <- c("barplot", "dotplot", "heatplot", "genefrequency", "upsetplot", "enrichmentmap", "treeplot")
  unlink(file.path(coll_path, paste0(prefix, "_", loop_plots, ".pdf")))

  # gene sets
  gene_sets <- msigdbr(species = "Homo sapiens", collection = collection, subcollection = subcollection) %>%
    dplyr::select(gs_name, gene_symbol)

  # ORA
  res <- enricher(gene          = df_sig$symbol,
                  pAdjustMethod = "BH",
                  TERM2GENE     = gene_sets,
                  universe      = background_genes,
                  minGSSize     = min_geneset_size,
                  maxGSSize     = 500,
                  pvalueCutoff  = padj_cutoff)

  # results table (all significant pathways)
  res_df_sig <- as.data.frame(res) %>%
    mutate(minuslog10padj = -log10(p.adjust))
  write.csv(res_df_sig, file.path(coll_path, paste0(prefix, "_resclusterprofiler.csv")), row.names = FALSE)

  # filters for plotting: significance, then keywords
  res_filtered <- res
  res_filtered@result <- res@result %>%
    filter(p.adjust <= padj_cutoff, qvalue <= res@qvalueCutoff)
  if (!is.na(keywords)) {
    res_filtered@result <- res_filtered@result %>%
      filter(str_detect(Description, keywords))
  }

  n_significant <- nrow(res_df_sig)
  n_plotted     <- nrow(res_filtered@result)
  cat("Gene sets:", length(unique(gene_sets$gs_name)),
      "| significant pathways:", n_significant,
      "| after keyword filter:", n_plotted, "\n")

  if (n_plotted == 0) {
    message("  No pathways to plot for ", label)
    return(tibble(label, n_significant, n_plotted))
  }

  title    <- paste(label, "Enrichment\n", name_of_comparison)
  pdf_file <- function(plot_name) file.path(coll_path, paste0(prefix, "_", plot_name, ".pdf"))

  # figures
  save_plot(function() barplot(res_filtered, showCategory = 15, title = title) + my_theme,
            pdf_file("barplot"), width = 10, height = 8)

  save_plot(function() dotplot(res_filtered, showCategory = 20, title = title,
                               x = "GeneRatio", color = "p.adjust", size = "Count") + my_theme,
            pdf_file("dotplot"), width = 10, height = 8)

  save_plot(function() heatplot(res_filtered, showCategory = 10, foldChange = gene_fc) +
              theme(axis.text.x = element_text(size = 4, angle = 90, hjust = 1, vjust = 0.5),
                    axis.text.y = element_text(size = 6)),
            pdf_file("heatplot"), width = 14, height = 8)

  save_plot(function() plot_gene_frequency(res_filtered,
                                           title = paste(label, "- most frequent genes in filtered pathways\n", name_of_comparison)),
            pdf_file("genefrequency"), width = 10, height = 6)

  save_plot(function() upsetplot(res_filtered, n = 15) +
              theme(axis.text.y = element_text(size = 6, face = "plain"),
                    axis.text.x = element_text(size = 8)),
            pdf_file("upsetplot"), width = 14, height = 8)

  # enrichment map and treeplot need pathways to compare, so they are skipped with fewer than 3
  if (n_plotted >= 3) {
    enrichres_sim <- pairwise_termsim(res_filtered)

    save_plot(function() emapplot(enrichres_sim, showCategory = 40,
                                  label_format = 25, node_label_size = 2),
              pdf_file("enrichmentmap"), width = 14, height = 12)

    save_plot(function() treeplot(enrichres_sim,
                                  label_format = 25,
                                  fontsize_tiplab = 2,
                                  fontsize_cladelab = 2,
                                  nCluster = min(5, n_plotted - 1), # fewer clusters when there are few pathways
                                  extend = 0.01,
                                  cladelab_offset = 6,
                                  hexpand = 0.7) +
                theme(legend.text  = element_text(size = 8),
                      legend.title = element_text(size = 9)),
              pdf_file("treeplot"), width = 20, height = 8)
  } else {
    message("  Skipped enrichment map and treeplot for ", label, ": fewer than 3 pathways")
  }

  tibble(label, n_significant, n_plotted)
}
```

### Run the loop

```{r}
loop_summary <- pmap_dfr(collections, run_collection) # runs run_collection() once per row of collections

loop_summary
```

```{r, results="hide"}
#significant DEG
write.csv(df_sig, file.path(out_path, paste0(name_of_comparison, '_degs.csv')), row.names = FALSE)

writeLines(capture.output(sessionInfo()),
           file.path(out_path, 'clusterProfiler_loop_sessionInfo.txt'))
```