---
title: "Overrepresentation Analysis with clusterProfiler"
format: html
editor: visual
---

### Install BiocManager, which is used to install and manage Bioconductor packages, if you don't have it already

```{r, results="hide"}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
```

### Once BiocManager is installed, you can use it to install clusterProfiler, enrichplot

```{r, results="hide"}
# Bioconductor packages
bioc_packages <- c("clusterProfiler", "enrichplot", "ggupset", "msigdbr")

BiocManager::install(setdiff(bioc_packages, rownames(installed.packages())), ask = FALSE)


# CRAN packages
cran_packages <- c("tidyverse", "readxl", "circlize", "pheatmap", "RColorBrewer")

install.packages(setdiff(cran_packages, rownames(installed.packages())))
```

### Load the required libraries

```{r, results="hide"}
# Loading relevant libraries 
library(tidyverse)        # data manipulation and visualization (ggplot2, dplyr, tidyr, etc.)
library(RColorBrewer)     # color palettes for plots
library(pheatmap)         # heatmaps with clustering
library(clusterProfiler)  # enrichment analysis (ORA and GSEA)
library(enrichplot)       # visualization of enrichment analysis results
library(msigdbr)          # molecular signatures database (pathways, biological functions, etc.)
library(readxl)           # reading Excel files into R
library(circlize)         # circular/ chord diagrams visualizations
```

### Set the path to your folder

```{r, results="hide"}
## Set relevant paths
base_path <- getwd() # get the current working directory as the root path
make_path <- function(folder) {
  path <- file.path(base_path, folder) # build the full path by appending folder name to base path
  if (!dir.exists(path)) dir.create(path, recursive = TRUE) # create the folder if it doesn't exist
  path                                      # return the path
}

in_path  <- make_path("input") # create and store path to the input folder
out_path <- make_path("output") # create and store path to the output folder
```

### Upload the results of the DESeq2 analysis to the input folder

```{r}
#df <- read_excel(file.path(in_path, 'AD_vs_ctrol_male'))
df <- read_csv(file.path(in_path, "AD_vs_ctrol_male.csv"))
genes_in_data <- df$symbol # extract the gene symbols column from the dataframe as a character vector
head(df)
```

### Define the DEGs cutoffs (e.g., log2FoldChange, padj)

```{r}

#Cleaning column names just in case
#colnames(df) <- trimws(colnames(df)) # when reading from Excel files, column names come in with hidden spaces- trimws cleans that up.

df_sig <- df %>%
  filter(!is.na(padj), !is.na(log2FoldChange)) %>% # remove rows with missing padj or log2FoldChange values
  mutate(diffexpressed = case_when( # create column diffexpress and classfies according to log2FC value
    log2FoldChange > 0.5 & padj < 0.05 ~ "UP",
    log2FoldChange < -0.5 & padj < 0.05 ~ "DOWN",
    TRUE ~ "NO" # all remaining genes: not significantly differentially expressed
  )) %>%
  filter(diffexpressed != "NO") %>% # keep only significantly UP or DOWN regulated genes
  distinct(symbol, .keep_all = TRUE) # remove duplicate gene symbols, keeping all columns


# Create UP and DOWN subsets from df_sig
df_up   <- df_sig %>% filter(diffexpressed == "UP")
df_down <- df_sig %>% filter(diffexpressed == "DOWN")


# Report, sanity check
cat("Found", nrow(df_sig), "DEGs\n")
cat("UP:  ", sum(df_sig$diffexpressed == "UP"), "\n")
cat("DOWN:", sum(df_sig$diffexpressed == "DOWN"), "\n")
```

```{r}
head(df_sig)
```

### Prepare background gene list

```{r}
# # extracting all unique, non-missing gene symbols from the full dataset to use as the statistical background (universe) for ORA
background_genes <- unique(na.omit(df$symbol))

cat("Number of background genes:", length(background_genes), "\n")
```

## Molecular signatures database

### To see available collections

```{r}
# more about msigdbr https://igordot.github.io/msigdbr/articles/msigdbr-intro.html
# more about Molecular Signatures Database https://www.gsea-msigdb.org/gsea/msigdb/index.jsp 
msigdbr_collections()
```

### To see available species collections

```{r}
msigdbr_species()
```

### Define the gene sets or molecular signatures

```{r}
name_of_gene_set <- 'C5' #simple character variable used as a label of collection
gene_sets_df <- msigdbr(species = 'Homo sapiens', collection = 'C5', subcollection = 'GO:BP')
#gene_sets_df is a tibble containing all gene sets from the specified MSigDB 
head(gene_sets_df)
```

### Convert gene sets to clusterProfiler format

```{r}
gene_sets <- gene_sets_df %>%
  dplyr::select(gs_name, gene_symbol) ## keeps only the gene set name and gene symbol columns, dropping all other metadata columns

cat(paste("Loaded", length(unique(gene_sets$gs_name)), "gene sets\n"))
```

```{r}
head(gene_sets)
```

## Running clusterProfiler

### Settings the variables

```{r}
name_of_comparison  <- 'AD_vs_ctrol_male' ## label for the comparison
padj_cutoff         <- 0.05 # significance threshold for enriched pathways
min_geneset_size    <- 7 # # minimum number of DEGs that must overlap with a pathway to be reported
filename_prefix     <- paste0(name_of_comparison, '_', name_of_gene_set) # builds a reusable file name prefix by combining the comparison and gene set labels # e.g. → male_AD_vs_male_ctrol_C2
```

### Run clusterProfiler Overrepresentation Analysis

We will use the enricher(), which is the universal generic function used to perform Over-Representation Analysis (ORA). It allows to run a hypergeometric test on any organism, custom gene set, or non-standard pathway database

```{r}
res <- enricher(gene          = df_sig$symbol, #all DEGs or replace:  df_up$Symbol or df_down$Symbol
                pAdjustMethod = "BH", # correcting for multiple testing using Benjamini-Hochberg (FDR) method
                TERM2GENE     = gene_sets, # gene set defined above
                universe      = background_genes, #defined above
                minGSSize     = min_geneset_size,
                maxGSSize     = 500, # exclude very large gene sets as they tend to be too broad and non-specific
                pvalueCutoff  = padj_cutoff) # only pathways with adjusted p-value < 0.05

cat(paste("Total enriched pathways:", nrow(as.data.frame(res)), "\n"))
```

### Extract the enrichment results

```{r}
res_df <- as.data.frame(res)

res_df_sig <- res_df %>%
  mutate(minuslog10padj = -log10(p.adjust)) # add a new column converting adjusted p-values to -log10 scale for visualization

head(res_df_sig)
```

### Saving the results

```{r}
#significant pathways
write.csv(res_df_sig, file.path(out_path, paste0(filename_prefix, '_resclusterprofiler.csv')), row.names = FALSE)
#significant DEG
write.csv(df_sig,     file.path(out_path, paste0(name_of_comparison, '_degs.csv')),            row.names = FALSE)
```

## Visualizations

### Applying filters

-filter first for significance, standardized enrichment score, ..., then keywords

```{r}
res_filtered <- res

res_filtered@result <- res@result %>%
  filter(p.adjust < 0.05) %>%      # keep only significant
  filter(str_detect(Description, "SYNAP|NEURO|SIGNAL|MEMBRAN|RECEPT|TRANS"))  # then keyword filter
#adapt these keywords to your biological context

cat("Significant pathways before keyword filter:", sum(res@result$p.adjust < 0.05), "\n")
cat("Pathways after keyword filter:             ", nrow(res_filtered@result), "\n")
```

### Barplot arranged by counts

```{r}
#barplot() from clusterProfiler returns a ggplot object so it can be customized
my_theme <- theme(
  axis.text.y = element_text(size = 6),
  axis.text.x = element_text(size = 8),
  plot.title  = element_text(size = 12),
  legend.text = element_text(size = 8)
)


p1 <- suppressWarnings(
  barplot(res_filtered,
          showCategory = 15,
          title = paste(name_of_gene_set, "Enrichment\n", name_of_comparison))
) + my_theme  

print(p1)

ggsave(filename = file.path(out_path, paste0(filename_prefix, "_barplot.pdf")),
       plot = p1, width = 10, height = 8)
```

### Barplot arranged by q-values

```{r}
p1a <- suppressWarnings(
  mutate(res_filtered %>% filter(!is.na(qvalue)), 
       log10qvalue = -log(qvalue, base = 10)) %>%
  barplot(x = "log10qvalue", showCategory = 15,
          title = paste(name_of_gene_set, "Enrichment\n", name_of_comparison))
) + my_theme

print(p1a)

ggsave(filename = file.path(out_path, paste0(filename_prefix, "_barplot2.pdf")),
       plot = p1a, width = 10, height = 8)
```

### Dotplot

```{r}

p2 <- suppressWarnings(
  dotplot(res_filtered, showCategory = 20,
          title = paste(name_of_gene_set, "Enrichment\n", name_of_comparison),
          x = "GeneRatio", color = "p.adjust", size = "Count") #gene ratio is DEGs in pathways/total DEGs for enrichment
) + my_theme
print(p2)

ggsave(filename = file.path(out_path, paste0(filename_prefix, "_dotplot.pdf")),
       plot = p2, width = 10, height = 8)
```

### Heatplot

```{r}
p3 <- suppressWarnings((heatplot(res_filtered, showCategory = 10)))
p3 <- p3 +
  theme(
    axis.text.x = element_text(size = 4, angle = 90, hjust = 1, vjust = 0.5),  # gene names
    axis.text.y = element_text(size = 6)                                         # pathways names
  )
print(p3)

ggsave(filename = file.path(out_path, paste0(filename_prefix, "_heatplot.pdf")),
       plot = p3, width = 14, height = 8)
```

### Enrichment map plot

```{r}
#Pairwise similarity is computed using the Jaccard similarity scores (from 0 to 1) between pathways

enrichres_sim <- pairwise_termsim(res_filtered)   

# Plotting
p4 <- emapplot(enrichres_sim, 
               showCategory = 40,
               label_format = 25,
               node_label_size = 2)   
print(p4)

ggsave(filename = file.path(out_path, paste0(filename_prefix, "_enrichmentmap.pdf")),
       plot = p4, width = 14, height = 12)
```

### Upset plot

clusterProfiler uses ggplot-based upset, NOT the full UpSetR package For full upset plot, install UpSetR or ComplexUpset

```{r}
p5 <- upsetplot(res_filtered, n = 15)
p5 <- p5 +
  theme(axis.text.y = element_text(size = 6, face = "plain"),
        axis.text.x = element_text(size = 8))
print(p5)

ggsave(filename = file.path(out_path, paste0(filename_prefix, "_upsetplot.pdf")),
       plot = p5, width = 14, height = 8)
```

### Treeplot

```{r}
#Similar to Enrichment Map, pairwise similarity is computed using the Jaccard similarity scores (from 0 to 1), converted to distance (1-Jaccard) between pathways
# we are using the already computed enrichres_sim in the Enrichment map plot section

p6 <- treeplot(enrichres_sim,
               label_format = 25,
               fontsize_tiplab = 2,     # size of individual pathway labels (tips)
               fontsize_cladelab = 2,   # size of cluster/group labels
               nCluster = 5,            # adjust number of clusters if needed
               extend = 0.01,           # reduce colored cluster area (default 0.3)
               cladelab_offset = 6,  # bring cluster labels closer (default 1), increase to separate from pathways labels
               hexpand = 0.7)          # expand horizontal space for pathway labels
p6 <- p6 +
  theme(
    legend.text  = element_text(size = 8),  #  legend text
    legend.title = element_text(size = 9)   #  legend title
  )

print(p6)

ggsave(filename = file.path(out_path, paste0(filename_prefix, "_treeplot.pdf")),
       plot = p6, width = 20, height = 8)
```

### Chord plot or chord diagram

```{r}
# Extract gene-pathway associations from res
gene_pathway <- as.data.frame(res_filtered) %>%
  dplyr::select(Description, geneID) %>%
  dplyr::slice_head(n = 15) %>%              # top 15 pathways
  tidyr::separate_longer_delim(geneID, delim = "/")    # one row per gene

# Keep only genes appearing in multiple pathways (for meaningful chords)
gene_counts <- gene_pathway %>%
  dplyr::count(geneID) %>%
  dplyr::filter(n > 4)                       # genes in more than 4 pathway

gene_pathway_filtered <- gene_pathway %>%
  dplyr::filter(geneID %in% gene_counts$geneID)

# Create adjacency matrix
chord_matrix <- table(gene_pathway_filtered$Description,
                      gene_pathway_filtered$geneID)
chord_matrix <- as.matrix(chord_matrix)

# defining colors for pathways (one gradient)
n_pathways <- nrow(chord_matrix)
n_genes    <- ncol(chord_matrix)

pathway_colors <- colorRampPalette(c("#2166AC", "#92C5DE"))(n_pathways)  # blue gradient
names(pathway_colors) <- rownames(chord_matrix)   # assign to pathway names

# defining colors for genes (different gradient)
gene_colors <- colorRampPalette(c("#D6604D", "#FDDBC7"))(n_genes)  # red gradient
names(gene_colors) <- colnames(chord_matrix)      # assign to gene names

# combine into one named color vector
all_colors <- c(pathway_colors, gene_colors)

# Plot chord diagram
pdf(file.path(out_path, paste0(filename_prefix, "_chordplot.pdf")),
    width = 14, height = 14)

set.seed(42)  # for reproducible chord diagram layout
circos.clear()
chordDiagram(
  chord_matrix,
  transparency = 0.4,
  annotationTrack = "grid",
  grid.col = all_colors, 
  preAllocateTracks = list(track.height = 0.3)  # space for labels
)

# Add pathway labels
circos.trackPlotRegion(
  track.index = 1,
  panel.fun = function(x, y) {
    sector.name <- get.cell.meta.data("sector.index")
    xlim        <- get.cell.meta.data("xlim")
    circos.text(
      mean(xlim), 0.5,
      sector.name,
      facing    = "clockwise",
      niceFacing = TRUE,
      adj        = c(0, 0.5),
      cex        = 0.5          # adjust label size
    )
  },
  bg.border = NA
)

circos.clear()
dev.off()
```

```{r, results="hide"}
writeLines(capture.output(sessionInfo()),
           file.path(out_path, 'clusterProfiler_sessionInfo.txt'))
```
